---
title: "Focal Ranking Test: Results & Workflow"
author: "NE Adams"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: hide
    theme: cosmo
    df_print: paged
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.align = "center", out.width = "100%"
)

suppressPackageStartupMessages({
  library(tidyverse)
})

# --- Path handling -----------------------------------------------------------
# Works whether the report is knit from the repo root (working dir = root)
# or via rmarkdown::render("analysis/focal_ranking_test.Rmd") (working dir =
# analysis/). No dependency on `here`.
base <- if (dir.exists("output")) "." else ".."
p <- function(...) file.path(base, ...)

# Include a figure only if it exists (older runs may lack some diagnostics).
show_fig <- function(...) {
  path <- p(...)
  if (file.exists(path)) knitr::include_graphics(path) else invisible()
}

# Render a CSV as a table if it exists; silently skip if not.
show_table <- function(path, caption = NULL, digits = 3) {
  if (length(path) == 0 || is.na(path) || !file.exists(path)) return(invisible())
  df <- readr::read_csv(path, show_col_types = FALSE)
  knitr::kable(df, caption = caption, digits = digits)
}

# The per-size summary CSV has inconsistent names across runs
# (ranking_summary_statistics_10k.csv vs ranking_summary_statistics.csv).
# Resolve whichever exists.
find_summary <- function(size) {
  cand <- p("output", paste0("focal_test_", size),
            c(sprintf("ranking_summary_statistics_%s.csv", size),
              "ranking_summary_statistics.csv"))
  hit <- cand[file.exists(cand)]
  if (length(hit)) hit[1] else NA_character_
}
```

# Overview

This page reports the **focal (database-search) ranking test**: take one member
of a simulated pair as a query profile, drop its true relative into a large
database of unrelated individuals, score every database member against the query
by combined likelihood ratio, rank them, and ask **whether the true relative
lands in the top *N* (200 / 100 / 50 / 10)** — and how that changes with the
true relationship, the tested hypothesis, the loci set, and the size of the
database.

- Method & module details: [`README_focal_ranking_test.md`](../README_focal_ranking_test.md)
- Core simulation / LR pipeline: [`SimulationModules_summary.md`](../SimulationModules_summary.md)

Each existing pair is one focal replicate: **10,000 replicates per true
relationship** (10 chunks × 1,000 pairs). Databases tested so far: **10k, 20k,
100k** unrelated individuals.

## Study parameters

| Parameter | Value |
|-----------|-------|
| Unrelated database | Pooled ("all") population allele frequencies; tested at **10k, 20k, 100k** individuals |
| Focal replicates | **10,000 per true relationship**, drawn under "all" (pooled) frequencies |
| Relatives simulated per focal | 1 parent–child, 1 full sibling, 1 half sibling |
| Hypotheses tested | Parent–child **and** full-sibling (both tested for every pair) |
| Loci sets | Core 13 and Expanded 20 |
| Outcome | The true relative's rank, and whether it lands in the top 10 / 50 / 100 / 200 |

# Workflow

Heavy computation runs on the cluster (SLURM); this report reads only the small
figure and summary outputs. Steps 1–5 run once per database size; step 6 runs
once across sizes.

```bash
# 1. Generate the unrelated pool (once per N)
sbatch code/generate_unrelated_pool.sh all 20000

# 2. Build the manifest (pairs + combined_LR + this pool)
code/make_focal_ranking_manifest.sh --pool-size 20000 --max-chunk 10

# 3. Rank (array = manifest rows x CHUNKS_PER_FILE)
sbatch --array=1-3000 code/submit_focal_ranking.sh \
  output/focal_test_20k/manifest_all_N20000_<dt>.csv \
  10 100 output/focal_test_20k/chunks

# 4. Combine per-chunk outcomes
code/combine_ranking_outcomes.sh \
  output/focal_test_20k/chunks output/focal_test_20k/combined_all_N20000

# 5. Per-database figures + summary table
Rscript code/plot_ranking_summary.R \
  output/focal_test_20k/combined_all_N20000_outcomes.csv \
  output/focal_test_20k/figures

# 6. Cross-database comparison (once, over all sizes)
Rscript code/compare_ranking_across_databases.R --out output/focal_cross_db \
  output/focal_test_10k/combined_all_N10000_outcomes.csv \
  output/focal_test_20k/combined_all_N20000_outcomes.csv \
  output/focal_test_100k/combined_all_N100000_outcomes.csv
```

# Results across database sizes

The headline question — how recovery holds up as the database grows.

## Recovery curves

Each curve is the empirical CDF of the true relative's rank: the proportion of
replicates with rank ≤ *N*, for every *N*. Database size is the line colour;
the tested hypothesis is the line type; dotted verticals mark top 10/50/100/200.
(The solid parent-child-tested lines that sit flat then jump near *N* ≈ database/2
are the tied-at-zero block — under a mismatched hypothesis most candidates share
no allele at some locus, collapse to LR = 0, and receive one averaged rank.)

```{r}
show_fig("output", "focal_cross_db", "figures", "recovery_curves_across_databases.png")
```

## Recovery at each threshold vs. database size

The same result read at the operational cutoffs, with Wilson 95% intervals.

```{r}
show_fig("output", "focal_cross_db", "figures", "recovery_by_threshold_across_databases.png")
```

## Top-200 recovery by database size

```{r}
rt_path <- p("output", "focal_cross_db", "recovery_by_threshold_across_databases.csv")
if (file.exists(rt_path)) {
  readr::read_csv(rt_path, show_col_types = FALSE) %>%
    filter(threshold == "Top 200") %>%
    select(true_relationship, tested_relationship, loci_set, db_size, prop) %>%
    pivot_wider(names_from = db_size, values_from = prop) %>%
    knitr::kable(digits = 3,
                 caption = "Proportion of true relatives recovered in the top 200")
}
```

## Rank summary

```{r}
show_table(p("output", "focal_cross_db", "rank_summary_across_databases.csv"),
           caption = "Median / mean / min / max rank of the true relative, per condition and database size")
```

# Results per database size {.tabset}

Per-size figures: recovery in top *N*, the full rank distribution, and (for runs
that recorded it) the size of the tied-at-zero group.

## 10k

```{r}
show_fig("output", "focal_test_10k", "figures", "ranking_prop_in_top_n.png")
show_fig("output", "focal_test_10k", "figures", "ranking_rank_distribution.png")
show_fig("output", "focal_test_10k", "figures", "ranking_tied_group_size.png")
```

```{r}
show_table(find_summary("10k"), caption = "10k database — summary statistics")
```

*Note: the 10k run predates the tied-group-size diagnostic, so that panel is
absent here.*

## 20k

```{r}
show_fig("output", "focal_test_20k", "figures", "ranking_prop_in_top_n.png")
show_fig("output", "focal_test_20k", "figures", "ranking_rank_distribution.png")
show_fig("output", "focal_test_20k", "figures", "ranking_tied_group_size.png")
```

```{r}
show_table(find_summary("20k"), caption = "20k database — summary statistics")
```

## 100k

```{r}
show_fig("output", "focal_test_100k", "figures", "ranking_prop_in_top_n.png")
show_fig("output", "focal_test_100k", "figures", "ranking_rank_distribution.png")
show_fig("output", "focal_test_100k", "figures", "ranking_tied_group_size.png")
```

```{r}
show_table(find_summary("100k"), caption = "100k database — summary statistics")
```

# Key findings

- **Parent–child and full-sibling relatives under the matched hypothesis are recovered ~98–100% in the top 200, nearly independent of database size.** A close relative's LR dominates any unrelated candidate. The one mild exception is full siblings at Core 13, which slip from 99.6% (10k) to 97.8% (100k).
- **Half-siblings are recoverable but graded, and recovery degrades as the database grows** — most steeply at fewer loci (Core 13 falls ~19 points across 10k→100k vs Expanded 20's ~11) and at tighter thresholds. This is the false-association-scales-with-database-size effect familial search is concerned with.
- **Testing the wrong hypothesis collapses recovery.** True full- and half-siblings scored under the parent-child hypothesis hit a no-sharing locus, zero out, and join a large tied-at-zero block (median tied group ≈ the whole database) — so top-*N* recovery, not median rank, is the interpretable metric there.
- **More loci help under the correct hypothesis, but hurt under the wrong one.** Expanded 20 beats Core 13 for every matched/appropriate case (most visibly half-siblings: 83% vs 62% in the top 200 at 100k). But under the mismatched parent-child hypothesis the ranking reverses — Expanded 20 recovers *fewer* true relatives (full-sib true: 4.0% vs 13.8%; half-sib true: 0.1% vs 1.2%), because each extra locus is another chance for a true relative to hit a no-sharing locus and zero out.

# Reproducing this report

```bash
# 1. Pull the small outputs (figures + summary CSVs) down from the cluster:
bash code/pull_focal_outputs.sh <user@cluster> <remote_repo>/output

# 2. Render (produces a self-contained analysis/focal_ranking_test.html):
Rscript -e 'rmarkdown::render("analysis/focal_ranking_test.Rmd")'
```

```{r}
sessionInfo()
```
